---
title: "02-DataManip"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

#Load Directories and Libraries

```{r load}

library(tidyverse)
library(reshape)
library(INLA)
library(corrplot)
#library(rgdal)
library(sf)
library(ggmap)
library(leaflet)
library(mapview)
library(rnaturalearth)

library("ggmap") 
register_stadiamaps("1cdeb586-3f5a-4ef9-813b-02375efa9f21")


out.dir <- paste("Output/")
dat.dir <- paste("Data/")
```

Anuran data are downloaded from the [CWMP](https://www.greatlakeswetlands.org/Home.vbhtml) website in `.xlsx` format. There are multiple tabs per spreadsheet, some which contain the metadata and others the raw data (points and obs), which are saved in the `Data` folder with a .csv

The site metadata files was provided separately by Doug Tozer. If a new files is needed, Doug or another member of the CWMP team should be contacted directly: dtozer\@birdscanada.org

What we will want to load into RStudio is the raw data, which are save in `.csv` format and in the `Data` directory in this R Project folder. If you need to update this analysis, simply re-save the raw data with the same file names in the `Data` folder.

#Load data and join tables

```{r data}

#Load data
bpoint<-read.csv("Data/CWMP_anuran_point.csv")
bobs<-read.csv("Data/CWMP_anuran_obs.csv")
site<-read.csv("Data/cwmp_master_site_table_sept2023.csv") #inlcude basin

#change site to site_id 
site<-site %>% mutate(site_id=site) %>% select(-site)

#Join tables 
frog_dat<-NULL
frog_dat<-left_join(bobs, bpoint, by=c("site_id", "crew_name", "date", "year", "sample", "point_id", "qa_done", "quar_flag", "longitude", "latitude", "coord_qual", "comments"))

```

Take a tour of the data to learn about its characteristics. Derive some summary stats to help you understand the data structure.

#Data exploration

```{r datatour}

#How many year has each sites been run? #How many points per site? 
sites<-frog_dat %>% group_by(site_id) %>% dplyr::summarize(siteyr=n_distinct(year), npoint=n_distinct(point_id)) 

mean(sites$siteyr) #2.
range(sites$siteyr) #1-11
mean(sites$npoint) #2.58
range(sites$npoint) #1-17

hist(sites$siteyr) #There are a lot of sites that have only been run once or twice
hist(sites$npoint)

#How many unique lat longs per points? 
latlong<-frog_dat %>% group_by(site_id, point_id) %>% dplyr::summarize(nlat=n_distinct(latitude), nlong=n_distinct(longitude), nyear=n_distinct(year))

#How may missing lat longs?
miss_ll<-frog_dat %>% group_by(site_id, point_id) %>% filter(is.na(latitude)) #2005 missing lat and long. Will want to rectify this issue. 

#Did you want to consider an observer effect?
length(unique(frog_dat$observer)) # 101 unique observers
obs<-frog_dat %>% group_by(observer) %>% dplyr::summarize(obsyr=n_distinct(year))
hist(obs$obsyr) #there are a lot of first year observers

#Do some observers survey multiple sites? 
length(unique(frog_dat$site_id)) # 747 unique sites
obs2<-frog_dat %>% group_by(observer) %>% dplyr::summarize(obssite=n_distinct(site_id))
hist(obs2$obssite)#there are a lot of observers that survey multiple sites. Likely because they are paid employees and not volunteers.

```

What this summary tells me is that a lat and long are not unique to the point_id within a site. It appears that a surveyor takes a new lat and long each year which is less efficient than working from fixed point count locations. We will need to assign a single lat&long per site_id (or point_id) for the analysis. Doug shared the master site list, which should have unique lat and long for each site ID.

There are a lot of observers that survey multiple point per year, but there are also a lot of observers that only survey for one year. We may consider including an `observer-site` random effect in the model. We might also want to have a first-year observer effect if we anticipate that first year observers are less proficient. Recall that surveyors and trained professionals and must pass a test to participate in the survey, so observer effects may be negligable.

#Data cleaning

There is a modifier `M` on some of the point_ids. Doug suggest these are site who's locations needed to be modified slightly within a given year. I will remove the modifier assuming they are close enough.

```{r mdrop}

frog_dat<-frog_dat %>% mutate(point=ifelse(point_id=="1M", "1", ifelse(point_id=="2M", "2", ifelse(point_id=="3M", "3", ifelse(point_id=="4M", "4", ifelse(point_id=="5M", "5", ifelse(point_id=="6M", "6", ifelse(point_id=="8M", "8", ifelse(point_id=="1MA", "1", point_id))))))))) %>% select(-point_id) 

frog_dat<-frog_dat %>% dplyr::rename(point_id=point)

```

Site 5755 is missing lat and longs. There doesn't appear to be any in the database for any year. Doug confirmed that this is a data entry error and they site should be 5735.

```{r sitefix}

frog_dat<-frog_dat %>% mutate(site_id=ifelse(site_id=="5755", "5735", site_id))

```

Now join to the master site file.

```{r masterjoin}

frog_dat<-frog_dat %>% select(-latitude, -longitude, -drw_flag, -drw_notes)

frog_dat$site_id<-as.integer(frog_dat$site_id)

frog_dat<-left_join(frog_dat, site, by="site_id")

```

#Sites surveyed per year summary

```{r sitesurvey}

site_sum<-frog_dat %>% select(site_id, year) %>% distinct() %>% mutate(value="x")
site_sum<-cast(site_sum, site_id~year, value="value")
nyear<-frog_dat %>% group_by(site_id) %>% summarise(nyears=n_distinct(year))
site_sum<-left_join(site_sum, nyear, by="site_id")
write.csv(site_sum, "Output/Summary_SiteSumYear.csv", row.names = FALSE)

```

#Types of sites surveyed per year (i.e., class)

```{r siteclass}

site_class<-frog_dat %>% group_by(class, year) %>% summarise(nsites=n_distinct(site_id))
site_class<-cast(site_class, class~year, value="nsites")
write.csv(site_class, "Output/Summary_ClassYear.csv", row.names=FALSE)

```

There seems to be good sample sized for each class type in each year.

#Raw species counts

```{r sp counts}

sp<-frog_dat %>% group_by(year, taxa_code) %>% summarize(sample_size=length(date))
sp<-cast(sp, year~taxa_code)

write.csv(sp, "Output/Summary_SpeciesYear.csv", row.names=FALSE)

```


#Assign each site to a BCR or Region

At first we were assigning each sample point to a BCR-Basin intersect to assign out of range, but then Doug created a unique puzzle to do this more accurately. The shape file is saved in the `Data` folder called `puzzelPieces.shp` but the near assignment was done in ArcGIS because some of the sample points fall outside the bounds of the .shp file. The old BCR code is still currently retained but could be deleted if found to be obsolete. 

```{r region}

#read region file 
region<-read.csv("Data/rangematrix.csv")
region<-region %>% dplyr::rename(site_id=site, region=label2)
frog_dat<-left_join(frog_dat, region, by=c("site_id"))

```

#Types of sites surveyed per year (i.e., class)

```{r siteclass}

site_basin<-frog_dat %>% group_by(basin, year) %>% summarise(nsites=n_distinct(site_id))
site_basin<-cast(site_basin, basin~year, value="nsites") %>% na.omit()

write.csv(site_basin, "Output/Summary_BasinYear.csv")

```

There seems to be good sample sized for each class type in each year. 

#Plot route locations and number of years surveyed 

```{r plot}

#Plot data point locations
plot_data<-frog_dat %>% group_by(site_id, lat, lon) %>% dplyr::summarize(siteyr=n_distinct(year)) 

#Create spatial data
plot <- frog_dat %>% select("lon", "lat")%>% distinct()
plot_sf <- st_as_sf(plot, coords = c("lon", "lat"), crs = 4326)

#Create the base layer map
map <- get_stadiamap(bbox = as.numeric(st_bbox(plot_sf)), zoom = 5)

#Create a new plot for each year
plot<-ggmap(map) + 
  geom_point(data = plot_data, aes(x = lon, y = lat, size=siteyr, colour=siteyr))
print(plot)

```

#Create column for raw occurance and an events matrix

```{r occurance}

frog_dat<-frog_dat %>% mutate(occ = ifelse(taxa_code=="(none)", 0, 1))

#site level zero-fill
events<-NULL
events<-frog_dat %>% select(site_id, point_id, year, lat, lon, class, basin, region, areaHa) %>% distinct(site_id, point_id, year, .keep_all = TRUE)

#Special events matrix to properly zero fill the occupancy frame. Notice this is nearly twice as long as the first events layer, which is expected. 
events2<-NULL
events2<-frog_dat %>% select(site_id, point_id, sample, year, region) %>% distinct()

```

#Create the speices list

```{r splist}

#Combine Eastern and Cope's Gray Tree Frog = TRFR
#Combine Boreal and Western Chorus Frog = CHFR
frog_dat<-frog_dat %>% mutate(ssp=ifelse(taxa_code %in% c("GRTR", "CGTR"), "TRFR", ifelse(taxa_code %in% c("CHFR", "BCFR"), "CHFR", taxa_code)))

#Remove Mink, Pickerel and Folwers Toad
frog_dat<- subset(frog_dat, !(taxa_code %in% c("MIFR", "PIFR", "FOTO", "(none)"))) 
sp.list<-unique(frog_dat$ssp)

```


#Species raw occupancy 

Here we explore the raw occupancy based on basin. This is raw occupancy based on three visits per site. 

```{r spsum}

#create dataframes and tables outside loop

master_basin<-NULL
master_basin<-events %>% select(region) %>% distinct() %>% na.omit()


for(n in 1:length(sp.list)){
  
#n<-1 #for testing the loop
  
  sp.data<-NULL #clear old data frame
  sp.data<-frog_dat %>% filter(ssp==sp.list[n])
  sp<-unique(sp.data$ssp)
  
#Retain only needed data
sp.data<-sp.data %>% select(occ, site_id, point_id, sample, year, lat, lon, class, region, areaHa) %>% group_by(site_id, point_id, sample, year, class, region) %>% dplyr::summarise(count=sum(occ)) %>% distinct()
  
#Merge with events layer to zero-fill
sp.data<-left_join(events2, sp.data, by=c("site_id", "point_id", "sample", "year", "region"))
  
#Zero fill the 'count' column
  sp.data$count[is.na(sp.data$count)] <- 0 
  sp.data$ssp<- sp.list[n]
  
#Let's summarize raw occupancy per region (this looks at each site-year combination to determine if it was occupied at least once). This dataframe may be most useful for establishing range. 
  sp.occ<-NULL
  sp.occ<-sp.data %>% group_by(year, site_id, region) %>% dplyr::summarize(sptot=sum(count)) %>%  mutate(occ=ifelse(sptot>=1, 1, 0))
  sp.occ<-sp.occ %>% group_by(region) %>% dplyr::summarize(tot=length(occ), raw=sum(occ), per_occ=(raw/tot)*100) %>% select(-tot, -raw)

colnames(sp.occ)[colnames(sp.occ) == "per_occ"] <- sp
master_basin<-left_join(master_basin, sp.occ, by="region")

}# end species loop

write.csv(master_basin, "Output/OccupancyPerRegion_full.csv", row.names = FALSE)

```

#Max occupancy per survey point

Creates the `CWMP_MaxOcc.csv` used for the analysis, which is a zero-filled data matrix. 

```{r range}

#Create table outside the species loop
maxcount<- as.data.frame(matrix(data = NA, nrow = 1, ncol = 12, byrow = FALSE, dimnames = NULL))
names(maxcount) <- c("site_id", "point_id", "year", "lat", "long", "class", "basin", "region", "areaHa", "maxocc", "maxcode", "taxa_code")
write.table(maxcount, file = paste(out.dir,"CWMP_MaxOcc.csv", sep=""), row.names = FALSE, append = FALSE, quote = FALSE, sep = ",")

#Read back in the master_basin file to assing range  
master_basin<-read.csv("Output/OccupancyPerRegion_full.csv")

for(n in 1:length(sp.list)){
  
#n<-1 #for testing the loop
  
  sp.data<-NULL #clear old data frame
  sp.data<-frog_dat %>% filter(ssp==sp.list[n])

#Retain only needed data, and max per point count visit
sp.data<-sp.data %>% select(occ, callcode_id, site_id, point_id, sample, year) %>% group_by(site_id, point_id, sample, year) %>% summarise(occ=max(occ), maxcode = max(callcode_id))


#Following Hohman et al. 2021, the dependent variable was the maximum occupancy in the three breeding-season point counts. Lets create this variable and write to a .csv 

maxocc<-sp.data %>% group_by(site_id, point_id, year)  %>% filter(maxcode==max(maxcode)) %>% select(-sample) %>% distinct()

#Merge with events layer to zero-fill
maxocc<-left_join(events, maxocc, by=c("site_id", "point_id", "year"))

#Zero fill the 'individuals' column
  maxocc$occ[is.na(maxocc$occ)] <- 0 
  maxocc$maxcode[is.na(maxocc$maxcode)] <- 0 
  maxocc$taxa_code<- sp.list[n]
  
#Create the range matrix using the per occ per basin > 10%

sp.range<-master_basin %>% select(region, sp.list[n])
colnames(sp.range)[colnames(sp.range) == sp.list[n]] <- "per_occ"

sp.range<-sp.range %>% mutate(range=ifelse(per_occ>=5, 1, 0)) %>% select(-per_occ)

#remove out of range observations and zeros
sp.data<-left_join(maxocc, sp.range, by="region")

#drop range == 0 
sp.data<-sp.data %>% filter(range>=1) %>% select(-range)
  
write.table(sp.data, file = paste(out.dir,"CWMP_MaxOcc.csv", sep=""), row.names = FALSE, append = TRUE, col.names=FALSE, quote = FALSE, sep = ",")  


} #end loop

```




